---
title: "session11 - homework"
format: html
editor: visual
date: 2025-07-17
author: Timo Roettger
execute:
  error: false
  warning: false
  message: false
  cache: true
bibliography: ../resources/bibliography.bib
---

# Preamble: Loading packages and configuration

```{r}
#| label: data_and_libraries
#| echo: false

# just run this code chunk
# function to ignoring the setting of the relative path below when knitting
run_if_not_knitting <- function(expr) {
  if (!isTRUE(getOption("knitr.in.progress"))) {
    eval(expr)
  }
}

# nifty code using the pacman package
# it checks if the packages specified below are installed, if not, they will be installed, if yes, they will be loaded
if (!require("pacman")) install.packages("pacman")
pacman::p_load(rstudioapi, tidyverse, tidybayes, brms, ggdist, bayestestR, modelr)

# set the current working directory to the one where this file is
run_if_not_knitting(current_working_dir <- dirname(rstudioapi::getActiveDocumentContext()$path))
run_if_not_knitting(setwd(current_working_dir))

# options to increase efficiency of brms models (optional)
options(brms.backend = "cmdstanr")
options(mc.cores = parallel::detectCores())

```

# Introduction

We will be analyzing the data from @roettger2014assessing.

```{r}
#| label: preprocessing

# Load our datasets
incNeutr <- read_csv("../data/E1_production.csv") |> 
  # we will focus only on the relevant columns
  select(subject, voicing, item_pair, usable, vowel_dur) |> 
  # and only on rows the authors deemed usable
  filter(usable == "usable") |> 
  select(!usable)

```

The variables:

`subject`: unique identifier for participants (n = 16)

`voicing`: underlying voicing status of stop (`voiced` vs. `voiceless`)

`item_pair`: unique identifier for pseudo word minimal pairs (n = 24)

`vowel_dur`: vowel duration in ms

In this study, we recorded people producing words such as Rad and Rat in German and were interested to see whether these words phonetically differ. The reason why this is interesting is that, according to German phonology, these words *should* be identical, i.e. he voicing distinction between /d/ and /t/ should have been *neutralized.* However, previous literature indicates that this neutralization is phonetically *incomplete.* For example, the vowel preceding a (de)voiced stop such /d/ is slightly longer preceding a voiceless stop such as /t/.

The researchers hypothesize that the vowel duration is meaningfully longer preceding voiced stops than preceding voiceless stops. They model this relationship using a linear mixed effects model that accounts for both speaker and word variation as well as by-speaker and by-word variation for the critical predictor voicing.

```{r}
#| label: model_formula

# store model formula
mdl_formula = bf(vowel_dur ~ voicing +
                   # random intercepts and slopes for subject and item_pair
                   (1 + voicing | subject) +
                   (1 + voicing | item_pair))

```

```{r}
#| label: spec-priors

# some weakly informative priors 
my_priors <- c(prior(normal(150,50), class = 'Intercept'),
               # weakly informative prior for the voicing difference (very diffuse)
               prior(normal(0, 30), class = 'b', coef = 'voicingvoiceless'),
               # weakly informative prior for all sources of variance (very diffuse)
               prior(cauchy(0,100), class = 'sd'),
               # weakly informative prior for residual variance (very diffuse)
               prior(cauchy(0,100), class = 'sigma'),
               # correlations
               prior(lkj(1), class = 'cor'))

```

```{r}
#| label: run-model

# run model
xmdl <- brm(mdl_formula,
            family = gaussian(),
            prior = my_priors,
            file = "../models/in_sesoi",
            data = incNeutr,
            seed = 42)

# I also store the prior-only model 
xmdl_prior <- brm(mdl_formula,
            family = gaussian(),
            prior = my_priors,
            sample_prior = "only",
            file = "../models/in_sesoi_priors",
            data = incNeutr,
            seed = 42)

```

# Exercises

## (a) Plot the posterior distribution

Explore the posterior distribution for the difference between voiced and voiceless stops. Interpret it.

```{r}

# YOUR ANSWERS HERE
# Interpretation: The posterior distribution for the effect of voicing on vowel duration is centered at −8.54 ms (SD = 2.27 ms), with a 95% HDI spanning [−13.17, −4.13] ms. The distribution is unimodal, relatively narrow, and symmetric, indicating a precise and well-constrained estimate. Crucially, the entire credible interval excludes zero — there is virtually no posterior mass near the null value — which provides strong evidence that vowels are systematically shorter before voiceless stops than before voiced stops. This result is consistent with the hypothesis of phonetically incomplete neutralization. Although German final devoicing is expected to render word-final stop contrasts phonologically neutralized, speakers systematically preserve a residual phonetic cue in the form of slightly longer vowels before historically voiced stops. The posterior distribution reflects this: the effect is not only credibly different from zero but is estimated with enough precision to confidently characterize its direction and approximate magnitude.
posterior_draws <- as_draws_df(xmdl)

# Plot the posterior
ggplot(posterior_draws, aes(x = b_voicingvoiceless)) +
  stat_halfeye(
    fill = "steelblue", alpha = 0.7,
    point_interval = mean_hdi,
    .width = c(0.89, 0.95)   # 89% and 95% credible intervals
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red", linewidth = 0.8) +
  labs(
    title = "Posterior distribution: Effect of voicing on vowel duration",
    subtitle = "Coefficient: voiceless vs. voiced (reference)",
    x = "Difference in vowel duration (ms)\n[voiceless − voiced]",
    y = "Density"
  ) +
  theme_minimal(base_size = 13)
posterior_summary(xmdl, variable = "b_voicingvoiceless")
```

## (b) Define a SESOI

Define a SESOI and justify it. Can you find an applied justification for a SESOI. Or a theoretical justification? Are there justifications grounded in measurement? What about justifications based on previous research? Define it and formalize it as a Region of Practical Equivalence as we have done before. Use the `bayesfactor_rope()` function to calculate the Bayes factor for and against the posteriors being inside the ROPE.

```{r}

# YOUR ANSWER HERE
rope <- c(-10,10)
rope_result <- bayesfactor_rope(posterior=xmdl,
                                null= rope, 
                                parameter= 'b_voicingvoiceless')
print(rope_result)
```

## (c) Check robustness of your analysis

How robust is your conclusion against different prior choices?

```{r}

# YOUR ANSWER HERE
#To assess robustness, the model was re-run under a skeptical prior (Normal(0, 5)) and a diffuse prior (Normal(0, 100)), and results were compared against the original prior (Normal(0, 30)).The ROPE Bayes Factors tell a revealing story about prior sensitivity. Under the original prior, BF = 4.10 indicates moderate evidence that the effect falls outside the ROPE — the voicing effect is practically meaningful. Under the skeptical prior, BF = 15.20 actually increases to strong evidence outside the ROPE — perhaps counterintuitively, the tighter prior forces the posterior into a narrower region that sits more cleanly outside the default ROPE boundaries. Under the diffuse prior however, BF = 0.942 — essentially 1 — indicating complete ambiguity, with evidence neither for nor against practical significance.This pattern reveals something important: the conclusion about practical significance is sensitive to prior choice, even though the direction of the effect is stable across all three models. With a diffuse prior the data alone are not sufficient to confidently establish that the effect exceeds the ROPE threshold — the posterior spreads too widely. The original and skeptical priors, by contributing some regularizing information, produce more decisive conclusions.Overall, the robustness check supports the qualitative finding of a negative voicing effect across all prior specifications. However, the degree of practical significance depends on prior choice, which should be acknowledged as a limitation — particularly given the relatively small sample of 16 speakers. The most defensible conclusion is that the incomplete neutralization effect is real and directionally consistent, but whether it crosses the threshold of practical relevance remains somewhat uncertain and prior-dependent.
priors_skeptical <- c(
  prior(normal(150, 50), class = 'Intercept'),
  prior(normal(0, 5), class = 'b', coef = 'voicingvoiceless'),  # tight: skeptical of large effects
  prior(cauchy(0, 100), class = 'sd'),
  prior(cauchy(0, 100), class = 'sigma'),
  prior(lkj(1), class = 'cor')
)

xmdl_skeptical <- brm(mdl_formula,
                      family = gaussian(),
                      prior = priors_skeptical,
                      file = "../models/in_sesoi_skeptical",
                      data = incNeutr,
                      seed = 42)

priors_diffuse <- c(
  prior(normal(150, 100), class = 'Intercept'),
  prior(normal(0, 100), class = 'b', coef = 'voicingvoiceless'),
  prior(cauchy(0, 100), class = 'sd'),
  prior(cauchy(0, 100), class = 'sigma'),
  prior(lkj(1), class = 'cor')
)

xmdl_diffuse <- brm(mdl_formula,
                    family = gaussian(),
                    prior = priors_diffuse,
                    file = "../models/in_sesoi_diffuse",
                    data = incNeutr,
                    seed = 42)

draws_original  <- as_draws_df(xmdl) |> mutate(model = "Original (sd=30)")
draws_skeptical <- as_draws_df(xmdl_skeptical) |> mutate(model = "Skeptical (sd=5)")
draws_diffuse   <- as_draws_df(xmdl_diffuse) |> mutate(model = "Diffuse (sd=100)")



rope_original  <- bayesfactor_rope(xmdl,           range = c(-10, 10), parameters = "b_voicingvoiceless")
rope_skeptical <- bayesfactor_rope(xmdl_skeptical, range = c(-10, 10), parameters = "b_voicingvoiceless")
rope_diffuse   <- bayesfactor_rope(xmdl_diffuse,   range = c(-10, 10), parameters = "b_voicingvoiceless")

print(rope_original)
print(rope_skeptical)
print(rope_diffuse)

```

# References

<!-- References will auto-populate in the refs div below -->

::: {#refs}
:::
